Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-08 paper/ms.tex
add rasterization option...
because google chrome is bad at rendering things

(literally firefox does it.  adobe acrobat reader is fine.  preview is
fine.  just chrome)

\newcommand{\mstype}{letter}

@@

% dip-counting pipeline
\newcommand{\nuniqdipflagged}{{368}} % up-to-date; get_lgb_list_count.py

\input{vals_cqv_table_statistics}

@@

\received{---}
\revised{---}
\accepted{---}
\shorttitle{Four Years of Complex Periodic Variables}

@@

% primary authors
\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

@@

The sharp features of CPV light curves can have durations as short as
5\% of the rotation period ($P_{\rm rot}$), which is too short to be
caused by starspots rotating into and out of view.  Starspots produce
flux variations with characteristic timescales of $P_{\rm rot}$ and
$0.5\,P_{\rm rot}$.  With finely-tuned viewing geometries, starspots
can produce dip durations as short as $\approx$$0.2\,P_{\rm rot}$, but
in such cases, limb darkening causes the dip amplitudes to be smaller
than the observed amplitudes of $\sim$1\% (see
\citealt{2017AJ....153..152S}, Figures 37-41).  Thus, a
``starspot-only'' scenario can be ruled out for many CPVs
\citep{2017AJ....153..152S,2019ApJ...876..127Z,2021MNRAS.500.1366K}.
Given that many CPVs cannot be explained by starspots alone, and
working under the assumption that all CPVs share the same basic
physical scenario, we discard the ``starspot-only'' scenario.
Instead, the correct explanation probably involves spatially
concentrated circumstellar material
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.

Figure~\ref{fig:f1} illustrates two proposed models for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm c} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep[e.g.][]{2017AJ....153..152S,2017MNRAS.471L.145F,2023MNRAS.518.4734S}.
The second scenario invokes ``prominences'', long-lived condensations
of cool, dense, marginally-ionized gas that are embedded within the
hotter corona and that corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical prominences are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that an
optically-thick ring obscures a narrow band of the stellar photosphere
\citep{2019ApJ...876..127Z}; hot spots passing behind such a ring
could produce sudden dips.  We disfavor this scenario for reasons
described in Appendix~\ref{app:ring}.

While the dust clump and gas prominence hypotheses both invoke
magnetically-entrained material, the two pictures differ in the origin
and the composition of the occulting material.   ``Dust clumps''
invoke opacity from dust, which would need to be collisionally charged
\citep{2023MNRAS.518.4734S}; such dust might be sourced from a
low-mass debris disk.  ``Gas prominences'' invoke opacity from
partially-ionized gas, perhaps bound-free transitions in hydrogen or a
molecular opacity.  The gas might be sourced from a stellar wind.
Unambiguous evidence in support of either scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

@@

The average period, determined by measuring the PDM peak period over
each sector independently, was $\langle P \rangle = 18.5560$\,hr.  The
range between the maximum and minimum sector-specific periods was
measured to be about one minute.   However, a period shift of
$\pm$1\,minute leads to large phase drifts over the entire timespan of
observations.  One minute is $\approx$1/1000$^{\rm th}$ of a period,
and we have observed 1500 cycles.  By folding with a fine grid of
trial periods, we found that the choice $P=18.5611 \pm 0.0001$ causes
more of the features in the LP 12-502 light curve to maintain constant
phases over the entire dataset.

@@

% Author contributions statements
LGB conceived the project, performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses, and wrote the manuscript.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
ADU identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
JNW, SR, and LAH significantly improved the clarity of the manuscript.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.

>>>>

\newcommand{\mstype}{article}

@@

\newcommand{\nuniqdipflagged}{{368}} % up-to-date; get_lgb_list_count.py
\input{vals_cqv_table_statistics}

@@

\received{2023 September 8}
\revised{---}
\accepted{---}
\shorttitle{Four Years of Complex Periodic Variables}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

@@

The sharp features of CPV light curves can have durations as short as
5\% of the rotation period ($P_{\rm rot}$), which is too short to be
caused by starspots rotating into and out of view.  Starspots produce
flux variations with characteristic timescales of $P_{\rm rot}$ and
$0.5\,P_{\rm rot}$.  With finely-tuned viewing geometries, starspots
can produce dip durations as short as $\approx$$0.2\,P_{\rm rot}$, but
in such cases, limb darkening causes the dip amplitudes to be smaller
than the observed amplitudes of $\sim$1\% (see
\citealt{2017AJ....153..152S}, Figures 37-41).  Thus, a
``starspot-only'' scenario can be ruled out for many CPVs
\citep{2017AJ....153..152S,2019ApJ...876..127Z,2021MNRAS.500.1366K}.
Given that many CPVs cannot be explained by starspots alone, and
working under the assumption that all CPVs share the same basic
physical scenario, we discard the ``starspot-only'' model.
Instead, the correct explanation probably involves spatially
concentrated circumstellar material
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.

Figure~\ref{fig:f1} illustrates two proposed configurations for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm c} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep[e.g.][]{2017AJ....153..152S,2017MNRAS.471L.145F,2023MNRAS.518.4734S}.
The second scenario invokes ``prominences'', long-lived condensations
of cool, dense, marginally-ionized gas that are embedded within the
hotter corona and that corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical prominences are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that an
optically-thick ring obscures a narrow band of the stellar photosphere
\citep{2019ApJ...876..127Z}; hot spots passing behind such a ring
could produce sudden dips.  We disfavor this scenario for reasons
described in Appendix~\ref{app:ring}.

While the dust clump and gas prominence hypotheses both invoke
magnetically-entrained material, the two pictures differ in the origin
and the composition of the occulting material.   ``Dust clumps''
invoke opacity from dust, which would need to be collisionally charged
\citep{2023MNRAS.518.4734S}, and which might be sourced from a
low-mass debris disk.  ``Gas prominences'' invoke opacity from
partially-ionized gas, perhaps bound-free transitions in hydrogen or a
molecular opacity.  The gas might be sourced from a stellar wind.
Unambiguous evidence in support of either scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

@@

The average period, determined by measuring the PDM peak period over
each sector independently, was $\langle P \rangle = 18.5560$\,hr.  The
range between the maximum and minimum sector-specific periods was
measured to be about one minute.   However, a period shift of
$\pm$1\,minute leads to large phase drifts over the entire timespan of
observations.  One minute is $\approx$1/1000$^{\rm th}$ of a period,
and we have observed 1500 cycles.  By folding with a fine grid of
trial periods, we found that the choice $P=18.5611 \pm 0.0001$\,hr causes
more of the features in the LP 12-502 light curve to maintain constant
phases over the entire dataset.

@@

% Author contributions statements
LGB conceived the project, performed the dip-counting search, light curve classification, cluster membership, SED, variability, and secondary-period analyses, and wrote the manuscript.
RJ and SR performed the Fourier-based analysis and contributed to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
ADU identified the magnetic B star connection.
LAH contributed to project design.
JNW, SR, and LAH significantly improved the clarity of the manuscript.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
